{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-07 thesis/education.tex
handle many todos in educ

\begin{table}[t]
	% TODO
	\begin{center}
		\small
	\begin{tabular}{r||r|r|r|r|r|r|r||r}
			& \bf S'15 & \bf F'15 & \bf S'16 & \bf F'16 & \bf S'17 & \bf F'17 & \bf S'18 & \bf Total \\
			\hline
			%			S15	F15	S16	F16	S17	F17	S18
			\bf Participating groups& 8	& 22	& 18	& 12	& 19	& 14	& 10	& 103	\\
			% nb not counting FPs
			Groups w/any bugs found	& 5	& 21	& 12	& 7	& 14	& 10	& 6	& 75	\\
			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
			Groups w/all ($\ne$0) bugs fixed
						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
			\hline
			% the 2 for f15 were pretty messy; i had to go back and get em by hand by grepping for NOTE
			% but this semester also had a LOT of seemingly true race NOTE mx lock inf loop,
			% so i counted only the det ones, of which there were 2 (both also in "fixed" category)
			% otherwise 1 race recursive mx and 6 race mutex loops and 1 race cvar loop were left as races.

@@

{\bf Statistical significance.}
\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
% TODO: run a combined A-D test for each of the two control pairs
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

% TODO: should i report the T.AD test statistic? i don't even know what it means lol but it's what the p value like, applies to..
\begin{figure}[h]
	\begin{center}
		\small
	\begin{tabular}{cr|r||r|r|r|r||cc}
		\multicolumn{3}{c||}{\bf Population} & \multicolumn{4}{c||}{\bf Distribution} & \multicolumn{2}{c}{\bf Significance} \\
		\bf project & \bf group & $N$ & \bf min & \bf mean & \bf max & \bf stddev & \bf $p$ vs users & \bf cutoff\\
		\hline
			& users		& 103	& 42	& 74.8	& 95	& 10.3	& --	& --	\\
		P2	& non-users	& 74	& 38	& 70.9	& 96	& 12.5	& 0.036	& 0.05	\\
			& pre-study	& 86	& 43	& 73.7	& 93	& 9.7	& 0.671	& 0.05	\\
			\hline
			& users		& 98	& 65	& 83.6	& 95	& 6.2	& --	& --	\\
		P3	& non-users	& 63	& 45	& 80.7	& 93	& 8.2	& 0.034	& 0.05	\\
			& pre-study	& 79	& 60	& 81.6	& 95	& 7.6	& 0.145	& 0.05	\\
	\end{tabular}
	\end{center}
	\caption{Detailed statistics from student grade distributions.}
	\label{tab:table-of-ze-studence}
\end{figure}

@@

\item
{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\cref{sec:410-secrecy}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
would speak more to the impact of Landslide's preemption traces
in helping to diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

@@

On the other hand, because the {\tt thr\_exit()} bug typically manifests as a use-after-free,
the official tests must stress the thread library until the memory in question gets re-allocated
in just the right interleaving pattern to cause corruption that leads to a visible crash.
Landslide, meanwhile, detects this error immediately upon access (\cref{sec:landslide-valgrind-mode}),
with no need for complex corruption conditions.
Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
it is difficult to say whether the heap checking alone, together with unit or stress tests,
would have been sufficient,
or whether heap checking and model checking were both necessary in concert to help them.
To isolate the impact of model checking,
I simulated the students having access to a stand-alone heap checker
by checking only the first thread interleaving with Landslide
% TODO - more specific section reference - quicksand det/nondet drpps eval
(similarly to \cref{sec:quicksand-eval})
and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
Re-classifying these into the ``correct'' group,
the 11-10-0-16 distribution becomes
% fisher.test(rbind(c(11,10),c(3,13)))$p.value -->  0.0475343
11-10-3-13,
with a new $p$ value of 0.048:
still positively correlated, but no longer significant under the multiplicity-corrected cutoff.

@@

The survey responses were very positive:
% TODO: corroborate this with snapshot data?
students reported being able both to diagnose and to verify as fixed the vast majority of Landslide's reported bugs,
compared Landslide favourably to stress testing,
and found the experience worthwhile and worth recommending.
Regarding the 100 CPU-hours question in particular,
students could approach answering it in several different ways:
do the three students who answered 20/80 think stress testing is four times as good as Landslide at finding bugs,
or that stress testing requires four times as long to reach the same point of diminishing returns?
On the other extreme, one student said they would spend all 100 CPU-hours on Landslide,
which probably speaks more to their enthusiasm than to a careful attempt to maximize expected number of bugs found
% TODO: put a section reference to future work resource exhaustion / malloc failure injection discussion
(even the author themself recognizes stress testing's advantages for certain types of bugs, such as resource exhaustion).
Nevertheless, the responses' overall bias to spending at least half the CPU-time on Landslide
shows clearly that the students found the experience worthwhile.

@@

Comparing the survey response distributions between CMU versus PSU and U. Chicago (\Cref{fig:survey}),
CMU students tended to verify their bugfixes more often by re-running Landslide
and found the preemption traces easier to understand,
while PSU and U. Chicago students generally compared Landslide more favourably to stress testing
(CMU students comparatively preferred the more nuanced answer that it depends on the type of bug),
and reported more often that \revisionminor{it} helped them understand concurrency better.
%Considering that CMU's 15-410 demands substantial prerequisite concurrency experience
%(15-213 \cite{csapp} or equivalent)
%whereas PSU's \psuos is more towards the average between CMU's 15-213 and 15-410
%(tempering the difficulty of P2 to make it more accessible (\cref{sec:overview-psu})),
Considering the higher demand CMU's 15-410 makes for prerequisite concurrency experience
compared to PSU's relative tempering of P2's difficulty to make it more accessible (\cref{sec:pebbles}),
% TODO: represent u c in this sentents?
these trends seem to correlate with the different levels of preparation each course's students had,
showing that students of various skill levels
can each benefit from the experience in different ways.

% TODO

@@

% TODO - u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
% anything else?

@@

% TODO: is this ok to report without IRB approval? ask dave
I took no special measures to compensate for bias in gender, race, or being non-natively English-speaking
during recruitment or the survey.
Surveying students to measure any differences in these demographics between study participants and the overall class population
would have required mandatory survey participation,
and in turn, a more rigorous IRB approval process.
According only to my memory of students who attended the Landslide clinics
(\cref{sec:education-pebbles-recruiting}),
the racial diversity was roughly representative of the class at large,
and the proportion of women I perceived was in fact somewhat higher than the overall gender ratio.
More scientific analysis of such statistics was deemed beyond the study's scope for now.

>>>>

\begin{table}[t]
	\begin{center}
		\small
	\begin{tabular}{r||r|r|r|r|r|r|r||r}
			& \bf S'15 & \bf F'15 & \bf S'16 & \bf F'16 & \bf S'17 & \bf F'17 & \bf S'18 & \bf Total \\
			\hline
			%			S15	F15	S16	F16	S17	F17	S18
			\bf Participating groups& 8	& 22	& 18	& 12	& 19	& 14	& 10	& 103	\\
			% nb not counting FPs
			Groups w/any bugs found	& 5	& 21	& 12	& 7	& 14	& 10	& 6	& 75	\\
			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
			Groups w/all ($\ne$0) bugs fixed
						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
			\hline
			% the 2 for f15 were pretty messy; i had to go back and get em by hand by grepping for NOTE
			% but this semester also had a LOT of seemingly true race NOTE mx lock inf loop,
			% so i counted only the det ones, of which there were 2 (both also in "fixed" category)
			% otherwise 1 race recursive mx and 6 race mutex loops and 1 race cvar loop were left as races.

@@

{\bf Statistical significance.}
\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

\begin{figure}[h]
	\begin{center}
		\small
	\begin{tabular}{cr|r||r|r|r|r||cc}
		\multicolumn{3}{c||}{\bf Population} & \multicolumn{4}{c||}{\bf Distribution} & \multicolumn{2}{c}{\bf Significance} \\
		\bf project & \bf group & $N$ & \bf min & \bf mean & \bf max & \bf stddev & \bf $p$ vs users & \bf cutoff\\
		\hline
			& users		& 103	& 42	& 74.8	& 95	& 10.3	& --	& --	\\
		P2	& non-users	& 74	& 38	& 70.9	& 96	& 12.5	& 0.036	& 0.05	\\
			& pre-study	& 86	& 43	& 73.7	& 93	& 9.7	& 0.671	& 0.05	\\
			\hline
			& users		& 98	& 65	& 83.6	& 95	& 6.2	& --	& --	\\
		P3	& non-users	& 63	& 45	& 80.7	& 93	& 8.2	& 0.034	& 0.05	\\
			& pre-study	& 79	& 60	& 81.6	& 95	& 7.6	& 0.145	& 0.05	\\
	\end{tabular}
	\end{center}
	\caption{Detailed statistics from student grade distributions.}
	\label{tab:table-of-ze-studence}
\end{figure}

@@

\item
{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\cref{sec:410-secrecy}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
would speak more to the impact of Landslide's preemption traces
in helping to diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

@@

On the other hand, because the {\tt thr\_exit()} bug typically manifests as a use-after-free,
the official tests must stress the thread library until the memory in question gets re-allocated
in just the right interleaving pattern to cause corruption that leads to a visible crash.
Landslide, meanwhile, detects this error immediately upon access (\cref{sec:landslide-valgrind-mode}),
with no need for complex corruption conditions.
Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
it is difficult to say whether the heap checking alone, together with unit or stress tests,
would have been sufficient,
or whether heap checking and model checking were both necessary in concert to help them.
To isolate the impact of model checking,
I simulated the students having access to a stand-alone heap checker
by checking only the first thread interleaving with Landslide
(similarly to \cref{sec:quicksand-eval-nondets})
and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
Re-classifying these into the ``correct'' group,
the 11-10-0-16 distribution becomes
% fisher.test(rbind(c(11,10),c(3,13)))$p.value -->  0.0475343
11-10-3-13,
with a new $p$ value of 0.048:
still positively correlated, but no longer significant under the multiplicity-corrected cutoff.

@@

The survey responses were very positive:
% corroborate this with snapshot data? -- haha, never
students reported being able both to diagnose and to verify as fixed the vast majority of Landslide's reported bugs,
compared Landslide favourably to stress testing,
and found the experience worthwhile and worth recommending.
Regarding the 100 CPU-hours question in particular,
students could approach answering it in several different ways:
do the three students who answered 20/80 think stress testing is four times as good as Landslide at finding bugs,
or that stress testing requires four times as long to reach the same point of diminishing returns?
On the other extreme, one student said they would spend all 100 CPU-hours on Landslide,
which probably speaks more to their enthusiasm than to a careful attempt to maximize expected number of bugs found
(even the author themself recognizes stress testing's advantages for certain types of bugs,
such as resource exhaustion; see \cref{sec:warpzone-verif}).
Nevertheless, the responses' overall bias to spending at least half the CPU-time on Landslide
shows clearly that the students found the experience worthwhile.

@@

Comparing the survey response distributions between CMU versus PSU and U. Chicago (\Cref{fig:survey}),
CMU students tended to verify their bugfixes more often by re-running Landslide
and found the preemption traces easier to understand,
while PSU and U. Chicago students generally compared Landslide more favourably to stress testing
(CMU students comparatively preferred the more nuanced answer that it depends on the type of bug),
and reported more often that \revisionminor{it} helped them understand concurrency better.
%Considering that CMU's 15-410 demands substantial prerequisite concurrency experience
%(15-213 \cite{csapp} or equivalent)
%whereas PSU's \psuos is more towards the average between CMU's 15-213 and 15-410
%(tempering the difficulty of P2 to make it more accessible (\cref{sec:overview-psu})),
Considering the higher demand CMU's 15-410 makes for prerequisite concurrency experience
compared to PSU's relative tempering of P2's difficulty to make it more accessible (\cref{sec:pebbles}),
these trends seem to correlate with the different levels of preparation each course's students had,
showing that students of various skill levels
can each benefit from the experience in different ways.

@@

% u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
% ^ haha, never
% anything else?

@@

I took no special measures to compensate for bias in gender, race, or being non-natively English-speaking
during recruitment or the survey.
Surveying students to measure any differences in these demographics between study participants and the overall class population
would have required mandatory survey participation,
and in turn, a more rigorous IRB approval process.
According only to my memory of students who attended the Landslide clinics
(\cref{sec:education-pebbles-recruiting}),
the racial diversity was roughly representative of the class at large,
and the proportion of women I perceived was in fact somewhat higher than the overall gender ratio.
More scientific analysis of such statistics was deemed beyond the study's scope for now.
